Modelling complex population structure using $F$-statistics and Principal Component Analysis

### 2021-11-15 fstats_pca_lewontin.tex
updated figures
submission before journal specific changes

\title{Modelling complex population structure using $F$-statistics and Principal Component Analysis}
\author{Benjamin M Peter}
\begin{document}
	\maketitle
\begin{abstract}
Human  genetic diversity is shaped by our complex history. Population genetic tools to understand this variation can broadly be classified into data-driven methods such as Principal Component Analysis (PCA), and model-based approaches such as $F$-statistics.
Here, I show that these two perspectives are closely related, and I derive explicit connections between the two approaches. I show that $F$-statistics have a simple geometrical interpretation in the context of PCA, and that orthogonal projections are the key concept to establish this link. I illustrate my results on two examples, one of local, and one of global human diversity. In both examples, I find that just using the first few PCs provides good population structure is sparse, and only a few components contribute to most statistics. Based on these results, I develop novel visualizations that allow for investigating specific hypotheses, checking the assumptions of more sophisticated models. My results extend $F$-statistics to non-discrete populations, moving towards more complete and less biased descriptions of human genetic variation. 
\end{abstract}
\section{Introduction}
As in most species, the genetic diversity of human populations has been influenced by our history and environment over the last several hundred thousand years \citep[e.g][]{cavalli-sforza1994,marciniak2017, reich2018a, nielsen2017, witt2022}. In turn, an important goal of population genetics is to use observed patterns of variation to  investigate and reconstruct the demographic and evolutionary history of our species \citep{schraiber2015, orlando2021}.

@@

\paragraph{F-statistics}
However, when the number of populations exceeds a few dozen,  even codifying reasonable population models can be prohibitively difficult. One approach is to pick a small set of ``representative'' samples, and restrict modeling to this subset \citep[e.g.][]{gravel2011, harney2021}. However, this has the drawback that a large proportion of the available data remains unused. An increasingly popular alternative approach, particularly in the analysis of human ancient DNA, is therefore to build up complex models from smaller building blocks based on the relationship between two, three or four populations.

The framework is based on a set of parameters called  $F$-statistics \textit{sensu} Patterson \citep{reich2009, patterson2012, peter2016}. Formal definition will be given in the Theory section; but an informal motivation starts with the null model that populations are related as a tree, in which each $F$-statistic measures the length of a particular set of branches. \citep[Figure \ref{fig:geom};][]{semple2003, peter2016}.

In most applications, $F$-statistics are estimated from  data, and then used as tests of treeness. In particular, under the assumption of a tree, $F_3$ is restricted to be non-negative, and many $F_4$-statistics will be zero \citep{semple2003, patterson2012}, and data that violates these constraints is incompatible with a tree-like relationship between populations. The canonical alternative model is an admixture graph (or phylogenetic network) \citep{patterson2012, huson2010}, which is a tree which allows for additional edges reflecting gene flow (Figure \ref{fig:admix}A). However, admixture graphs are not the only plausible alternative model, and expected $F$-statistics can be calculated for a wide range of population genetic demographic models \citep{peter2016}.

@@

The most widely considered alternative model is an admixture graph \citep{patterson2012}, an example is given in Figure \ref{fig:admix}A. Here, the *typically unobserved) population $X_y$ is generated by a mixture of individuals from the ancestors of $X_2$ and $X_3$. Over time, genetic drift will change $X_y$ to $X_x$, which is the admixed population we observe. This will result in $F_4$-statistics that are non-zero, and, in some cases, in negative $F_3$-statistics \citep[exact conditions can be found in ][]{peter2016}.

\subsubsection{Geometric interpretation of $F$-statistics}
An implicit assumption in the development of $F$-statistics is that population lineages are mostly discrete, and that gene flow is rare. Recently, \cite{oteo-garcia2021} showed re-deriving $F$-statistics in a geometric framework, showing that these assumptions are not necessary. Specifically, they interpret the populations $X_i$ as points or vectors in the $S$-dimensional \emph{allele frequency space} $\mathbb{R}^S$. In this case, the $F$-statistics can be thought of as inner (or dot) products, and they showed that all properties and tests related to treeness can be derived in this larger space. In particular the $F$-statistics can be written as
\begin{subequations}
	\begin{align}
	&F_2(X_1, X_2) &=& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})^2
	&=& \frac{1}{S}\langle X_1 - X_2, X_1 - X_2 \rangle = \frac{1}{S}\normsq{X_1-  X_2}\\
	&F_3(X_1; X_2, X_3) &=& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})(x_{1l} - x_{3l}) &=& \frac{1}{S}\langle X_1 - X_2, X_1 - X_3 \rangle\\	
	&F_4(X_1, X_2; X_3, X_4) &=& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})(x_{3l} - x_{4l}) &=& \frac{1}{S}\langle X_1 - X_2, X_3 - X_4 \rangle	\text{,}
	\end{align}
\end{subequations}
where $\norm{\cdot}$ denotes the Euclidean norm and $\langle \cdot, \cdot \rangle$ denotes the dot product. Some elementary properties of the dot product between vectors $a, b, c$ that I will use later are
\begin{subequations}
	\begin{align}
	\langle a, b \rangle &= \sum_i a_ib_i\\
	\langle a, b \rangle &= \norm{a}\norm{b}\cos(\phi)\\
	\langle a, a \rangle &= \normsq{a}\\
	\langle a + c, b \rangle &= \langle a, b \rangle + \langle b, c \rangle,
	\end{align}
\end{subequations}
where $\phi$ is the angle between $a$ and $b$. The inner product is  closely related to vector projections
\begin{equation}
\vectorproj[b]{a} = \frac{\langle a , b\rangle}{\normsq{b}} b,\label{eq:proj}
\end{equation}
which is a vector colinear to $b$ whose length measures how much vector $a$ points in the direction of $b$. Thinking of $F$-statistics as projections also holds on trees: In e.g. a $F_4(X_1, X_4; X_2, X_3)$-statistic (Figure \ref{fig:admix}C), the internal branch is precisely the intersection of the paths from $X_1$ to $X_4$ and from $X_2$ and $X_3$. The external branches are independent populations, and so they are expected to drift orthogonally to each other.

@@

\subsection{Formal Definition of PCA}
PCA is a common way of summarizing genetic data, and so a large number of variations of PCA exist, e.g. in how SNPs are standardized, how missing data is treated or whether we use individuals or populations as units of analysis. The version of PCA I use here is set up such that the similarities to  $F$-statistics are maximized, and does \emph{not} reflect how PCA is most commonly applied to genome-scale human genetic variation data sets. In particular, I assume that a PCA is performed on unscaled, estimated population allele frequencies, whereas many applications of PCA are based on individual-level sample allele frequency, scaled by the estimated standard deviation of each SNP \citep{patterson2006}. The differences this causes will be addressed in the discussion.

Let us again assume we have allele frequency data as above, but let us now assume we aggregate the allele frequency vectors $X_i$ of many populations in a matrix $\MX$ whose entry $x_{il}$ reflects the allele frequency of the $i$-th population at the $l$-th genotype. If we have $S$ SNPs and $n$ populations, $\MX$ will have dimension $n \times S$. Since the allele frequencies are between zero and one, we can interpret each Population $X_i$
of $\MX$ as a point in $[0, 1]^S$, the allele frequency or \emph{data space}, which is a subset of $\mathbb{R}^S$.

One way PCA can be motivated is that it aims to find a $K$-dimensional subspace of the data space that retains most  variation in the data. $K$ is at most $n-1$, in which case the data is simply rotated. However, the  historical processes that generated genetic variation often result in \emph{low-rank} data \citep{engelhardt2010}, so that $K \ll n$ explains a substantial portion of the variation; for visualization $K=2$ is frequently used.

@@

\section{Material \& Methods}
The theory outlined in the previous section suggests that $F$-statistics have a geometric interpretation in PCA-space, which can be approximated on PCA plots. In the next section I explore this connection in detail, and illustrate it on two sample data sets that I briefly introduce here. Both are based on the analyses by \cite{lazaridis2014}. The data is from the Reich lab compendium data set (v44.3), downloaded from \url{https://reich.hms.harvard.edu/allen-ancient-dna-resource-aadr-downloadable-genotypes-present-day-and-ancient-dna-data}, using data on the ``Human Origins''-SNP set (597,573 SNPs). SNPs with missing data in any population are excluded. The code used to create all figures and analyses will be available on \url{https://github.com/BenjaminPeter/fstats_pca}.

@@

\paragraph{Western Eurasian data set}
This data set of 1,119 individuals from 62 populations contains  present-day individuals from the Eastern Mediterranean, Caucasus and Europe. It is motivated by the analysis of \cite{lazaridis2014}, who used it as a basis of comparison for ancient genetic analyses of Western Eurasian individuals, and PCAs based on similar sets of samples have been used in many other ancient DNA studies \citep[e.g.][]{lazaridis2016, haak2015}. Genetic differentiation in this region is low and closely mirrors geography \citep{novembre2008}. I thus speculate that gene flow between these populations is common \citep{ralph2013}, and a discrete model such as a tree or an admixture graph might be a rather poor reflection of this data. A file with all individuals used and their assigned population is given in \textbf{Supplementary File 2.}

\paragraph{Computing $F$-statistics and PCA}
All computations are performed in \texttt{R}. I use \texttt{admixtools 2.0.0} (\url{https://github.com/uqrmaie1/admixtools}) to compute $F$-statistics. To obtain a PC-decomposition, I first calculate all pairwise $F_2$-statistics, and then use equation \ref{eq:mds} and the \texttt{eigen} function to obtain the PCs. The right-hand side matrix of equation \ref{eq:mds} is supposed to have non-negative eigenvalues (i.e. $-\MC\MF\MC$ is positive-semidefinite). However as $F_2$-statistics are estimates, some eigenvalues might be slightly negative, which would lead to imaginary PCs. I avoid this by using the \texttt{nearPD}-function in R that ensures all eigenvalues have the correct sign.

@@

%\subsection{Geometry of $F$-statistics in PC-space}
The transformation from the previous section allows us to consider the geometry of $F$-statistics in PCA-space. The relationships we will discuss formally only hold if we use all PCs. However, the appeal of PCA is that frequently, only a very small number $K \ll n$ of PCS contain most information that is
relevant for population structure, in which case the geometric interpretations become very simple. Thus, throughout the schematic figures, I assume that two PCs are sufficient to characterize population structure. In the data applications I evaluate how deviations of this assumption my manifest themselves in PCA plots.

\subsection{$F_2$ in PC-space}
The $F_2$-statistic is an estimate of the squared allele-frequency distance between two
populations. On a tree (Figure \ref{fig:geom}A) this corresponds to the branch between two populations. In allele-frequency space, it corresponds to the squared Euclidean distance, and thus reflects the intuition that closely related populations  will fall close to each other on a PCA-plot, and have low pairwise $F_2$-statistics. However, since $F_2$ can be written as  a sum of squared (non-negative) terms for each PC (eq. \ref{eq:fpc}), the distance on a PCA-plot will always be an underestimate of the full $F_2$-distance. Thus, PCA might project two populations with high $F_2$-distance very close to each other, which would indicate that these particular PCs are not suitable to  understand and visualize the relationship between these particular populations. In converse, populations that are distant on a PCA-plot are guaranteed to also have a large $F_2$-distance.

@@

Thus, a natural question to ask is given two source populations $X_2$, $X_3$, can we use PCA to predict which populations might be considered admixed between them? One way to address this question is to consider the space for which $F_3$ is negative, i.e. 
\begin{eqnarray}
2 F_3(X_x; X_2, X_3) &=& 2\langle  X_x - X_2, X_x - X_3 \rangle \nonumber\\
      &=& \normsq{X_x - X_2} + \normsq{X_x - X_3}  - \normsq{X_2 - X_3} < 0 \label{eq:f3neg}\text{.}
\end{eqnarray}
By the Pythagorean theorem, $F_3 = 0 $ if and only if $X_2, X_3$ and $X_x$ form a right-angled triangle. The associated region where $F_3=0$ is a $n$-sphere (or a circle in two dimensions) with diameter $\overline{X_2X_3}$ (The overline denotes a line segment). $F_3$ is negative when the triangle is obtuse, i.e. $X_x$ could be considered admixed if it lies inside the $n$-ball with diameter $\overline{X_2X_3}$ (Figure \ref{fig:geom}B, Equation \ref{eq:f3sphere}).

@@

\paragraph{Example}
As an example, I visualize the admixture statistic $F_3(X; \text{Sardinian}, \text{Finnish})$, on the first two PCs of the Western Eurasian data set (Figure \ref{fig:f3}A). In this case, the projected $n$-ball (light gray) and circle based on two dimensions (dark gray) have similar sizes. However, several populations that appear inside the circles (e.g. Basque, Canary Islandersh) have, in fact, positive $F_3$-values, so they lie outside the $n$-ball. This reveals that the first two PCs do not capture all the genetic variation relevant for  European population structure.  Consequently, approximating $F_3$ by the first two or even ten PCs (Figure \ref{fig:f3}B) only gives a coarse approximation of $F_3$, and from Figure \ref{fig:f3}C we see that many higher PCs contribute to $F_3$ statistics.

@@

\begin{figure}[!ht]
	\includegraphics[width=.8\textwidth]{figures/outgroup_f3.pdf}
	\caption{\textbf{Outgroup-$F_3$-statistics}. Interpretation of outgroup $F_3$-statistic on a tree (A) and PCA-plot (B).  The red segment represents $F_3(X_O; X_U, X_3)$ and the dashed blue segment reflects $F_3(X_O; X_U, X_1)$ and $F_3(X_O; X_U, X_2)$, which have the same value. 
	}
	\label{fig:outgroupf3}
\end{figure}
\subsection{Outgroup-$F_3$-statistics as projections}
A common application of $F_3$-statistics is, given an unknown sample $X_U$, to find the most closely related population among a reference panel $(X_i)$ \citep{raghavan2014}. This is done  using an \emph{outgroup}-$F_3$-statistic $F_3(X_O; X_U, X_i)$, where $X_O$ is an outgroup. The reason an outgroup is introduced is to account for differences in sample times and additional drift in the reference populations (\ref{fig:outgroupf3}A) The outgroup-$F_3$-statistic $F_3(X_O; X_U, X_3)$ represents the branch length from $X_O$ to the common node between the three samples in the statistic, and the closer this node is to $X_U$, the longer the branch and hence the larger the $F_3$-statistic.

@@

On the right-hand-side terms only the $F_3$ term depends on the $X_i$. The fraction can be thought of as a normalizing constant,  which justifies the argument that the $F_3$-statistic and length of the projected vector are proportional to each other, and can thus be interpreted similarly. Thus, the outgroup-$F_3$-statistic is largest for whichever $X_i$ projects furthest along the axis from the outgroup to the unknown population; in the example in Figure \ref{fig:outgroupf3}B this is $X_3$.

\begin{figure}[!ht]
	\includegraphics[width=\textwidth]{figures/fig_data_world.pdf}	
	\caption{\textbf{PCA and $F$-statistics for the World data set} A: Visualization of Outgroup-$F_3$-statistic $F_3(\text{Mbuti}; \text{Sardinian}, X)$ on a PCA-biplot. The color of points correspond to the value of the $F_3$ statistic, with brighter yellows indicating higher values, i.e. higher similarity to Sardinians. The $F_3$-projection axis is given by a black line, the projection of populations onto this axis by thin gray lines. In the full data space, these projection are orthogonal to the axis. B: Projection along the axis Sardinian-Mbuti (X-axis), and PCA on residual of this projection (PC1 on Y-axis, PC2 as coloring). C: Approximation of $F_3(\text{Mbuti}; \text{Sardinian}, X)$ using the first two (blue) and first ten (red) PCs, respectively. D: Contributions of first ten PCs to all statistics of the form $F_3(\text{Mbuti}; \text{Sardinian}, X)$. E: Contributions of the first ten PCs to select $F_4$-statistics.  
	}	
	\label{fig:f4}
\end{figure}

@@

\paragraph{Example}
In Figure \ref{fig:f4}A, I use the World data set to visualize the outgroup-$F_3$-statistic $F_3(\text{Mbuti}; \text{Sardinian}, X_i)$, in i.e. a statistic that aims to find the population most closely related to Sardinian (a Mediterranean Island), assuming the Mbuti are an outgroup to all populations in the data set. On a PCA, we can interpret this $F_3$ statistic as the projection of the line segment from $\text{Mbuti}$ to population $X_i$ onto the line through Mbuti and Sardinians (black line). For each population, the projection is indicated with a grey line. In the full data space, this line is always orthogonal to the segment Mbuti-Sardinian, but on the plot (i.e. the subspace spanned by the first two PCs), this is not necessarily the case.  The coloring is based on the $F_3$-statistic calculated from all the data, with brighter values indicating higher $F_3$-statistics. In this case, the first two PCs approximate the $F_3$-statistic very well: Particularly the samples from East Asia, Siberia and the Americas (cluster in the top left of the plot) project very close to orthogonally, suggesting that most of the genetic variation relevant for this analysis is captured by these first two PCs.  We can quantify this and find that the first two PCs slightly underestimate the absolute value of $F_3$ (Figure \ref{fig:f3}C), but keep the relative ordering. I also find that many PCs, e.g. PCs 3-5, 7 and 10 have almost zero contribution to all $F_3$-statistics (Figure \ref{fig:f3}D), and PCs 6, 8 and 9 having a similar non-zero contribution for almost all statistics, likely because these PCs explain within-African variation.

>>>>

\title{Modelling complex population structure using $F$-statistics and Principal Component Analysis}
\author{Benjamin M Peter}
\begin{document}
	\maketitle
\begin{abstract}
Human  genetic diversity is shaped by our complex history. Data-driven methods such as Principal Component Analysis (PCA) are an important population genetic tool to understand this method. 
Here, I contrast PCA with a set of statistics motivated by trees ($F$-statistics). Here, I show that these two methods are closely related, and I derive explicit connections between the two approaches. I show that $F$-statistics have a simple geometrical interpretation in the context of PCA, and that orthogonal projections are the key concept to establish this link. I illustrate my results on two examples, one of local, and one of global human diversity. In both examples, I find that just using the first few PCs provides good population structure is sparse, and only a few components contribute to most statistics. Based on these results, I develop novel visualizations that allow for investigating specific hypotheses, checking the assumptions of more sophisticated models. My results extend $F$-statistics to non-discrete populations, moving towards more complete and less biased descriptions of human genetic variation. 
\end{abstract}
\section{Introduction}
As in most species, the genetic diversity of human populations has been influenced by our history and environment over the last several hundred thousand years \citep[e.g][]{cavalli-sforza1994,marciniak2017, reich2018a, nielsen2017, witt2022}. In turn, an important goal of population genetics is to use observed patterns of variation to  investigate and reconstruct the demographic and evolutionary history of our species \citep{schraiber2015, orlando2021}.

@@

%\paragraph{F-statistics}
However, when the number of populations exceeds a few dozen,  even codifying reasonable population models can be prohibitively difficult. One approach is to pick a small set of ``representative'' samples, and restrict modeling to this subset \citep[e.g.][]{gravel2011, harney2021}. However, this has the drawback that a large proportion of the available data remains unused. An increasingly popular alternative approach, particularly in the analysis of human ancient DNA, is therefore to build up complex models from smaller building blocks based on the relationship between two, three or four populations.

The framework is based on a set of parameters called  $F$-statistics \textit{sensu} Patterson \citep{reich2009, patterson2012, peter2016}. Formal definitions will be given in the Theory section; but an informal motivation starts with the null model that populations are related as a tree, in which each $F$-statistic measures the length of a particular set of branches. \citep[Figure \ref{fig:geom};][]{semple2003, peter2016}.

In most applications, $F$-statistics are estimated from  data, and then used as tests of treeness. In particular, under the assumption of a tree, $F_3$ is restricted to be non-negative, and many $F_4$-statistics will be zero \citep{semple2003, patterson2012}, and data that violates these constraints is incompatible with a tree-like relationship between populations. The canonical alternative model is an admixture graph (or phylogenetic network) \citep{patterson2012, huson2010}, which is a tree which allows for additional edges reflecting gene flow (Figure \ref{fig:admix}A). However, admixture graphs are not the only plausible alternative models, and expected $F$-statistics can be calculated for a wide range of population genetic demographic models \citep{peter2016}.

@@

The most widely considered alternative model is an admixture graph \citep{patterson2012}, an example is given in Figure \ref{fig:admix}A. Here, the (typically unobserved) population $X_y$ is generated by a mixture of individuals from the ancestors of $X_2$ and $X_3$. Over time, genetic drift will change $X_y$ to $X_x$, which is the admixed population we observe. This will result in $F_4$-statistics that are non-zero, and, in some cases, in negative $F_3$-statistics \citep[exact conditions can be found in ][]{peter2016}.

\subsubsection{Geometric interpretation of $F$-statistics}
An implicit assumption in the development of $F$-statistics is that population lineages are mostly discrete, and that gene flow is rare. Recently, \cite{oteo-garcia2021}  re-derived $F$-statistics in a geometric framework, showing that these assumptions are not necessary. Specifically, they interpret the populations $X_i$ as points or vectors in the $S$-dimensional \emph{allele frequency space} $\mathbb{R}^S$. In this case, the $F$-statistics can be thought of as inner (or dot) products, and they showed that all properties and tests related to treeness can be derived in this larger space. In particular the $F$-statistics can be written as
\begin{subequations}
	\begin{align}
	&F_2(X_1, X_2) &=& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})^2
	&=& \frac{1}{S}\langle X_1 - X_2, X_1 - X_2 \rangle = \frac{1}{S}\normsq{X_1-  X_2}\\
	&F_3(X_1; X_2, X_3) &=& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})(x_{1l} - x_{3l}) &=& \frac{1}{S}\langle X_1 - X_2, X_1 - X_3 \rangle\\	
	&F_4(X_1, X_2; X_3, X_4) &=& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})(x_{3l} - x_{4l}) &=& \frac{1}{S}\langle X_1 - X_2, X_3 - X_4 \rangle	\text{,}
	\end{align}
\end{subequations}
where $\norm{\cdot}$ denotes the Euclidean norm and $\langle \cdot, \cdot \rangle$ denotes the dot product. Some elementary properties of the dot product between vectors $a, b, c$ that I will use later are
\begin{subequations}
	\begin{align}
	\langle a, b \rangle &= \sum_i a_ib_i\\
	\langle a, b \rangle &= \norm{a}\norm{b}\cos(\phi)\\
	\langle a, a \rangle &= \normsq{a}\\
	\langle a + c, b \rangle &= \langle a, b \rangle + \langle b, c \rangle,
	\end{align}
\end{subequations}
where $\phi$ is the angle between $a$ and $b$. The inner product is  closely related to the vector projection
\begin{equation}
\vectorproj[b]{a} = \frac{\langle a , b\rangle}{\normsq{b}} b,\label{eq:proj}
\end{equation}
which is a vector colinear to $b$ whose length measures how much vector $a$ points in the direction of $b$. Thinking of $F$-statistics as projections also holds on trees: In e.g. a $F_4(X_1, X_4; X_2, X_3)$-statistic (Figure \ref{fig:geom}C), the internal branch is precisely the intersection of the paths from $X_1$ to $X_4$ and from $X_2$ and $X_3$. On trees, all disjoint paths are independent (i.e. orthogonal) from each other, and thus the external branches vanish under the projection.

@@

\subsection{Formal Definition of PCA}
PCA is a common way of summarizing genetic data, and so a large number of variations of PCA exist, e.g. in how SNPs are standardized, how missing data is treated, what algorithm is used, or whether we use individuals or populations as units of analysis. The version of PCA I use here is set up such that the similarities to $F$-statistics are maximized, and does \emph{not} reflect how PCA is most commonly applied to genome-scale human genetic variation data sets. In particular, I assume that a PCA is performed on unscaled, estimated population allele frequencies. In contrast, many applications of PCA are based on individual-level sample allele frequency, scaled by the estimated standard deviation of each SNP \citep{patterson2006}. The differences this causes will be addressed in the discussion.

Let us again assume we have allele frequency data as above, but let us now assume we aggregate the allele frequency vectors $X_i$ of many populations in a matrix $\MX$ whose entry $x_{il}$ reflects the allele frequency of the $i$-th population at the $l$-th genotype. If we have $S$ SNPs and $n$ populations, $\MX$ will have dimension $n \times S$. Since the allele frequencies are between zero and one, we can interpret each population $X_i$
of $\MX$ as a point in $\mathbb{R}^S$.

PCA allows us to approximate the points in the high-dimensional allele frequency space by a $K$-dimensional subspace of the  data. If all PCs are considered, $K=n-1$, in which case the data is simply rotated. However, the  historical processes that generated genetic variation often result in \emph{low-rank} data \citep{engelhardt2010}, so that $K \ll n$ explains a substantial portion of the variation; for visualization $K=2$ is frequently used.

@@

\section{Material \& Methods}
The theory outlined in the previous section suggests that $F$-statistics have a geometric interpretation in PCA-space, which can be approximated on PCA plots. In the next section I explore this connection in detail, and illustrate it on two sample data sets that I briefly introduce here. Both are based on the analyses by \cite{lazaridis2014}. The data is from the Reich lab compendium data set (v44.3), downloaded from \url{https://reich.hms.harvard.edu/allen-ancient-dna-resource-aadr-downloadable-genotypes-present-day-and-ancient-dna-data}, using data on the ``Human Origins''-SNP set (597,573 SNPs). SNPs with missing data in any population are excluded. The code used to write this paper, create all figures and analyses is  available on \url{https://github.com/BenjaminPeter/fstats_pca}.

@@

\paragraph{Western Eurasian data set}
This data set of 1,119 individuals from 62 populations contains  present-day individuals from the Eastern Mediterranean, Caucasus and Europe.  \cite{lazaridis2014} used this data set as a basis of comparison for ancient genetic analyses of Western Eurasian individuals, and PCAs based on similar sets of samples have been used in many other ancient DNA studies \citep[e.g.][]{lazaridis2016, haak2015}. Genetic differentiation in this region is low and closely mirrors geography \citep{novembre2008}. I thus speculate that gene flow between these populations is common \citep{ralph2013}, and a discrete model such as a tree or an admixture graph might be a rather poor reflection of this data. A file with all individuals used and their assigned population is given in \textbf{Supplementary File 2.}

\paragraph{Computing $F$-statistics and PCA}
All computations are performed in \texttt{R}. I use \texttt{admixtools 2.0.0} (\url{https://github.com/uqrmaie1/admixtools}) to compute $F$-statistics. To obtain a PC-decomposition, I first calculate all pairwise $F_2$-statistics, and then use equation \ref{eq:mds} and the \texttt{eigen} function to obtain the PCs. The right-hand side matrix of equation \ref{eq:mds} is supposed to have non-negative eigenvalues (i.e. $-\MC\MF\MC$ is positive-semidefinite). However as $F_2$-statistics are estimates, some eigenvalues might be slightly negative, which would lead to imaginary PCs. I avoid this by setting all negative eigenvalues to zero.

@@

%\subsection{Geometry of $F$-statistics in PC-space}
The transformation from the previous section allows us to consider the geometry of $F$-statistics in PCA-space. The relationships we will discuss formally only hold if we use all PCs. However, the appeal of PCA is that frequently, only a very small number $K \ll n$ of PCS contain most information that is
relevant for population structure, in which case the geometric interpretations become very simple. Thus, throughout the schematic figures, I assume that two PCs are sufficient to characterize population structure. In the data applications I evaluate how deviations of this assumption may manifest themselves in PCA plots.

\subsection{$F_2$ in PC-space}
The $F_2$-statistic is an estimate of the squared allele-frequency distance between two
populations. On a tree (Figure \ref{fig:geom}A) this corresponds to the branch between two populations (Figure \ref{fig:geom}E). In allele-frequency space, it corresponds to the squared Euclidean distance, and thus reflects the intuition that closely related populations  will fall close to each other on a PCA-plot, and have low pairwise $F_2$-statistics. However, since $F_2$ can be written as  a sum of squared (non-negative) terms for each PC (eq. \ref{eq:fpc}), the distance on a PCA-plot will always be an underestimate of the full $F_2$-distance. Thus, PCA might project two populations with high $F_2$-distance very close to each other, which would indicate that these particular PCs are not suitable to  understand and visualize the relationship between these particular populations, and likely more PCs need to be investigated to understand how these populations are related to each other. In converse, populations that are distant on a PCA-plot are guaranteed to also have a large $F_2$-distance.

@@

Thus, a natural question to ask is given two source populations $X_2$, $X_3$, can we use PCA to predict which populations might have negative $F_3$-statistics? This condition can be written as  
\begin{eqnarray}
2 F_3(X_x; X_2, X_3) &=& 2\langle  X_x - X_2, X_x - X_3 \rangle \nonumber\\
      &=& \normsq{X_x - X_2} + \normsq{X_x - X_3}  - \normsq{X_2 - X_3} < 0 \label{eq:f3neg}\text{.}
\end{eqnarray}
By the Pythagorean theorem, $F_3 = 0 $ if and only if $X_2, X_3$ and $X_x$ form a right-angled triangle. The associated region where $F_3=0$ is a $n$-sphere (or a circle in two dimensions) with diameter $\overline{X_2X_3}$ (The overline denotes a line segment). $F_3$ is negative when the triangle is obtuse, i.e. $X_x$ could be considered admixed if it lies inside the $n$-ball with diameter $\overline{X_2X_3}$ (Figure \ref{fig:geom}B, Equation \ref{eq:f3sphere}).

@@

\paragraph{Example}
As an example, I visualize the admixture statistic $F_3(X; \text{Sardinian}, \text{Finnish})$, on the first two PCs of the Western Eurasian data set (Figure \ref{fig:f3}A). In this case, the projected $n$-ball (light gray) and circle based on two dimensions (dark gray) have similar sizes. However, several populations that appear inside the circles (e.g. Basque, Canary Islanders) have, in fact, positive $F_3$-values, so they lie outside the $n$-ball. This reveals that the first two PCs do not capture all the genetic variation relevant for  European population structure.  Consequently, approximating $F_3$ by the first two or even ten PCs (Figure \ref{fig:f3}B) only gives a coarse approximation of $F_3$, and from Figure \ref{fig:f3}C we see that many higher PCs contribute to $F_3$ statistics.

@@

\begin{figure}[!ht]
	\includegraphics[width=.6\textwidth]{figures/outgroup_f3.pdf}
	\caption{\textbf{Outgroup-$F_3$-statistics}. Interpretation of outgroup $F_3$-statistic on a tree (A) and PCA-plot (B).  The red segment represents $F_3(X_O; X_U, X_3)$ and the dashed blue segment reflects $F_3(X_O; X_U, X_1)$ and $F_3(X_O; X_U, X_2)$, which have the same value. 
	}
	\label{fig:outgroupf3}
\end{figure}
\subsection{Outgroup-$F_3$-statistics as projections}
A common application of $F_3$-statistics is, given an unknown sample $X_U$, to find the most closely related population among a reference panel $(X_i)$ \citep{raghavan2014}. This is done  using an \emph{outgroup}-$F_3$-statistic $F_3(X_O; X_U, X_i)$, where $X_O$ is an outgroup. The reason an outgroup is introduced is to account for differences in sample times and additional drift in the reference populations (Figure \ref{fig:outgroupf3}A). The outgroup-$F_3$-statistic $F_3(X_O; X_U, X_3)$ represents the branch length from $X_O$ to the common node between the three samples in the statistic, and the closer this node is to $X_U$, the longer the branch and hence the larger the $F_3$-statistic.

@@

Of the right-hand-side terms, only the $F_3$ term depends on the $X_i$. The fraction can be thought of as a normalizing constant,  so the $F_3$-statistic is proportional to the length of  the projected vector, and they can thus be interpreted similarly. This means, that the outgroup-$F_3$-statistic is largest for whichever $X_i$ projects furthest along the axis from the outgroup to the unknown population; in Figure \ref{fig:outgroupf3} this is $X_3$.

\begin{figure}[!ht]
	\includegraphics[width=\textwidth]{figures/fig_data_world.pdf}	
	\caption{\textbf{PCA and $F$-statistics for the World data set} A: Visualization of Outgroup-$F_3$-statistic $F_3(\text{Mbuti}; \text{Sardinian}, X)$ on a PCA-biplot. The color of points correspond to the value of the $F_3$ statistic, with brighter yellows indicating higher values, i.e. higher similarity to Sardinians. The $F_3$-projection axis is given by a black line, the projection of populations onto this axis by thin gray lines. In the full allele frequency  space, these projection are orthogonal to the axis. B: Projection along the axis Sardinian-Mbuti (X-axis), and PCA on residual of this projection (PC1 on Y-axis, PC2 as coloring). C: Approximation of $F_3(\text{Mbuti}; \text{Sardinian}, X)$ using the first two (blue) and first ten (red) PCs, respectively. D: Contributions of first ten PCs to all statistics of the form $F_3(\text{Mbuti}; \text{Sardinian}, X)$. E: Contributions of the first ten PCs to select $F_4$-statistics.  
	}	
	\label{fig:f4}
\end{figure}

@@

\paragraph{Example}
In Figure \ref{fig:f4}A, I use the World data set to visualize the outgroup-$F_3$-statistic $F_3(\text{Mbuti}; \text{Sardinian}, X_i)$, i.e. a statistic that aims to find the population most closely related to Sardinian (a Mediterranean Island), assuming the Mbuti are an outgroup to all populations in the data set. On a PCA, we can interpret this $F_3$ statistic as the projection of the line segment from $\text{Mbuti}$ to population $X_i$ onto the line through Mbuti and Sardinians (black line). For each population, the projection is indicated with a grey line. In the full allele frequency space, this line is always orthogonal to the segment Mbuti-Sardinian, but on the plot (i.e. the subspace spanned by the first two PCs), this is not necessarily the case.  The coloring is based on the $F_3$-statistic calculated from all the data, with brighter values indicating higher $F_3$-statistics. In this case, the first two PCs approximate the $F_3$-statistic very well: Particularly the samples from East Asia, Siberia and the Americas (cluster in the top left of the plot) project very close to orthogonally, suggesting that most of the genetic variation relevant for this analysis is captured by these first two PCs.  We can quantify this and find that the first two PCs slightly underestimate the absolute value of $F_3$ (Figure \ref{fig:f3}C), but keep the relative ordering. I also find that many PCs, e.g. PCs 3-5, 7 and 10 have almost zero contribution to all $F_3$-statistics (Figure \ref{fig:f3}D), and PCs 6, 8 and 9 having a similar non-zero contribution for almost all statistics, likely because these PCs explain within-African variation.

\subsection{$F_4$-statistics as angles}
One interpretation of $F_4$ on PCA plots is similar to that of $F_3$; as a projection of one vector onto another, with the difference that now all four points may be distinct. $F_4$-statistics that correspond to a branch in a tree (as in Figure \ref{fig:geom}C), can be interpreted as being proportional to the length of a projected segment on a PCA plot (Figure \ref{fig:geom}G), again with the caveat that we need to scale it by a constant. If the $F_4$-statistic corresponds to a branch that does not exist in the tree, i.e it is a test statistic (Figure \ref{fig:geom}D), then, from the tree-interpretation, we expect $F_4(X_1, X_2; X_3, X_4) = 0$ which implies that the vectors $X_1 - X_2$ and $X_3 - X_4$ are orthogonal to each other, i.e. that $X_1$ and $X_2$  map to the same point on the projection axis $\overline{X_3X_4}$ (Figure \ref{fig:geom}H). In the case of an admixture graph, this is no longer the case: Both population $X_y$ and $X_x$ in Figure \ref{fig:admix}D do \emph{not} map to the same point as $X_1$ or $X_2$ do, implying that statistics of the form $F_4(X_1, X_x; X_3, X_4) \neq 0$.
